Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels

### 2023-07-11 manuscript_english.tex
update tex

Our results reveal that fitness prevalence corresponds to the fixation
index of \emph{S. aberti} populations with respect to a source
population. Populations closer to the historical endemic area present a
higher genetic diversity and a lower \(F_{st}\) value. Finally, this
study allows us to add a biogeographic explanation to the results
obtained with population genetic methods and to generate maps of this
structure as tools to support conservation with a perspective that
integrates both population genetics and historical patterns of species
distribution.

@@

@@

For a given region \(\mathbf{G}\) in te geography, at any point in time,
we have a set of existing environmental conditions
\(\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}\).
We define \(\vec{e}_i =(e_1, e_2, ..., e_v )_i\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables \(\mathbf{E}(\mathbf{G}, t)\). Then
\(R_0(\vec{e}_i)\) is a function that relates fitness to environment,
that is, for a species with non-overlapping generations, the
population's net reproductive rate \(R_0\) is a measure of fitness that
depends on environmental conditions. If we take the set of places in
environmental space where these values are greater than a threshold, we
can define a volume in multidimensional space such that \(R(\vec{e})\)
is always positive and greater than this threshold, that is
\(\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}\). i. e. the set of
environmental conditions for which values of fitness are capable of
sustaining populations. We call
\(\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)\). This is a way of
expressing the fundamental niche as a function of environmental
conditions. So we can map each point of geography to ambient space. and,
for each site in the geography there will correspond a site in the
environmental space with different environmental conditions and
different growth rates. However all sites within the region defined by
\(\mathbf{N}_F\) will have viable populations and positive rates
depending on time because we assume that the environment is dynamic and
each geographic point is static, so for each site in geographic space
this growth rate depends on the environment and therefore is different
at each time: \(R_0 = R_0(\vec{e}(t)) \}\). Thus locally the growth rate
will be related to the local population size as long as it is within the
boundary defined by \(\mathbf{N}_F\).

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{all_figures/figure_1.png}
\caption{Hypothetical niche in environmental space. An arbitrary point
moves within the ambient space entering and exiting the volume (niche)
where the intrinsic growth rate is positive.}
\end{figure}

@@

Moreover, the effective population can be estimated as the harmonic mean
of the population in a demographic time series (Karlin 1968) for
\(\tau\) generations
\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\).
We substitute the harmonic mean into the above equation, considering
that although the number of generations is not exactly equal to the time
of the fixation index, they are expected to be very close.

\[
F_{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_j} 
\]

@@

We call this harmonic mean of environmental suitability
\emph{Suitability prevalence index }(\(S\)) given between \(0\) and
\(1\), since it reflects the rate of population growth as a function of
environment over a period of time. Thus we have a way to check that the
loss of heterozygosity (through the fixation index \(F_{st}\)) is
inversely proportional to the fitness prevalence for an \(i\) site in
the geography:

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic mean of fitness \emph{Suitability prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
\(F_{st}\) is lower are in areas where the prevalence over time is
higher, i.e.~environmental conditions remain suitable for a significant
period of time and therefore no effects of genetic drift of populations
have occurred. Conversely, it is expected that in populations with a
high fixation index value, fitness will not have prevailed over the same
time interval even though conditions currently exist and populations are
present.

@@

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from @Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_st\) between this populations in matrix form. The
matrix was transformed into a table as pairs between populations with
index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, georeferences were
obtained for each population considered in the fixation index matrix.
From this geo-referencing the SPA value for each population was
extracted. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.

>>>>

@@

Our results reveal that suitability prevalence corresponds to the
fixation index of \emph{S. aberti} populations with respect to a source
population. Populations closer to the historical endemic area present a
higher genetic diversity and a lower \(F_{st}\) value. Finally, this
study allows us to add a biogeographic explanation to the results
obtained with population genetic methods and to generate maps of this
structure as tools to support conservation with a perspective that
integrates both population genetics and historical patterns of species
distribution.

@@

For a given region \(\mathbf{G}\) in te geography, at any point in time,
we have a set of existing environmental conditions
\(\mathbf{E}(\mathbf{G}, t) = \Big \{ \vec{e}_i; \mathbf{G}, t\Big \}\).
We define \(\vec{e}_i =(e_1, e_2, ..., e_v )_i\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables \(\mathbf{E}(\mathbf{G}, t)\). Then
\(R_0(\vec{e}_i)\) is a function that relates fitness to environment,
that is, for a species with non-overlapping generations, the
population's net reproductive rate \(R_0\) is a measure of fitness that
depends on environmental conditions. If we take the set of places in
environmental space where these values are greater than a threshold, we
can define a volume in multidimensional space such that \(R(\vec{e})\)
is always positive and greater than this threshold, that is
\(\mathbf{N}_F=\{\vec{e} | R_0(\vec{e}) \ge k \}\). i. e. the set of
environmental conditions for which values of fitness are capable of
sustaining populations. We call
\(\mathbf{N}_F \subseteq \mathbf{E}(\mathbf{G}, t)\). This is a way of
expressing the fundamental niche as a function of environmental
conditions. So we can correspong each point of geography to
environmental space, with different environmental conditions and
different growth rates. However all sites within the region defined by
\(\mathbf{N}_F\) will have viable populations and positive rates
depending on time because we assume that the environment is dynamic and
each geographic point is static, so for each site in geographic space
this growth rate depends on the environment and therefore is different
at each time: \(R_0 = R_0(\vec{e}(t)) \}\). Thus locally the growth rate
will be related to the local population size as long as it is within the
boundary defined by \(\mathbf{N}_F\).

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,height=\textheight]{all_figures/figure_1.png}
\caption{Hypothetical niche in environmental space. An arbitrary point
moves within the environmental space entering and exiting the volume
(niche) where the intrinsic growth rate is positive.}
\end{figure}

@@

Moreover, the effective population can be estimated as the harmonic mean
of the population in a demographic time series (Karlin 1968) for
\(\tau\) generations
\(\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)\)
for each \(i\) site in \(\mathbf{G}\). We substitute the harmonic mean
into the above equation, considering that although the number of
generations is not exactly equal to the time of the fixation index, they
are expected to be very close:

\[
F_{i}^{st} \approx \Big (kt\Big)  \frac{1}{\tau}\sum_{j=1}^{\tau}\frac{1}{N_{ij}} 
\]

@@

We call this harmonic mean of environmental suitability
\emph{Suitability prevalence index }(\(S\)) given between \(0\) and
\(1\), since it reflects the rate of population growth as a function of
environment over a period of time. Thus we have a way to check that the
loss of heterozygosity (through the fixation index \(F_{st}\)) is
inversely proportional to the suitability prevalence for an \(i\) site
in the geography:

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic \emph{Suitability Prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) since we expect that areas where the
\(F_{st}\) is lower are in areas where the prevalence over time is
higher, i.e.~environmental conditions remain suitable for a significant
period of time and therefore no effects of genetic drift of populations
have occurred. Conversely, it is expected that in populations with a
high fixation index value, fitness will not have prevailed over the same
time interval even though conditions currently exist and populations are
present.

@@

To test our hypothesis of genetic diversity as a function of
environmental suitability prevalence, from a set of genetically
structured populations, a matrix of fixation indices between populations
is considered. For this work we considerer the data from @Bono et al.
(2018). It consideres 10 populations along the known distribution and
provides values of \(F_{st}\) between this populations in matrix form.
The matrix was transformed into a table as pairs between populations
with index values between each pair. The values of the population with
itself, i.e.~\(0\), were included. Subsequently, georeferences were
obtained for each population considered in the fixation index matrix.
From this geo-referencing the SPA value for each population was
extracted. Finally, the SPA value was aligned in the table of fixation
index pairs. In total the table has 4 columns: source population, target
population, fixation index and SPA of the target population. This table
was separated by source population groups to generate a set of linear
models between SPA values and fixation indexes \(F_{st}\). An analysis
was performed for each source population with the hypothesis that the
correlation between SPA and fixation index exists and is non-zero if a
dispersal process has occurred from the historical endemic area of the
taxon.
